\documentclass[11pt,a4paper]{article}

\usepackage[margin=2.5cm]{geometry}
\usepackage{booktabs}
\usepackage{array}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{placeins}
\usepackage{amsmath,amssymb}

\title{Automatic SpLiCE-Guided Mitigation of Spurious Correlations in Self-Supervised Representations}
\author{Grigory Grechkin}
\date{\today}

\begin{document}
\maketitle

\section{Project Summary}

This project tests the hypothesis that concept-level information from a frozen
vision--language model can identify and weaken shortcut information learned by a
self-supervised encoder. SpLiCE is used to discover concepts associated with the
known spurious attribute after conditioning on the target label. The resulting
signal is then used in three distinct ways: (i) as an audit of dataset bias,
(ii) as guidance for SimCLR training through targeted augmentation or a
label-conditioned correlation regularizer, and (iii) as a direct sparse
SpLiCE concept-bottleneck (CBM) intervention baseline. The trained SSL encoder
remains a ResNet; the CLIP/SpLiCE model is frozen.

The central empirical prediction is not merely an increase in average accuracy.
If the intervention reduces shortcut reliance, worst-group (WG) validation and
test accuracy should improve relative to a matched SimCLR baseline, while a
probe trained to predict the spurious attribute should become less successful.
All hyperparameter selection is performed on validation data and the test split
is reserved for a final, preselected configuration.

\section{Brief Description of SpLiCE}

SpLiCE, Sparse Linear Concept Embeddings, decomposes dense CLIP image embeddings
into sparse nonnegative weights over a vocabulary of human-interpretable text
concepts. Instead of treating a CLIP image feature as an opaque vector, SpLiCE
approximates it as a linear combination of text-concept embeddings. This makes
it possible to assign each image a score for selected concepts, such as concepts
related to the Waterbirds background. In this project SpLiCE is used as a frozen
helper model, not as the SSL encoder being trained.

The current configuration uses an \texttt{open\_clip:ViT-B-32} backbone with
\texttt{laion2b\_s34b\_b79k} weights, the LAION vocabulary of size 10\,000, and
an $\ell_1$ penalty of 0.25. SpLiCE returns a sparse nonnegative concept vector.
After automatic dataset-level selection of ten coordinates, their full
per-image vector is retained for regularization. A scalar maximum is used only
for routing images to the strong-augmentation branch. The augmentation and
correlation objectives are extensions developed in this project rather than
methods claimed by the original SpLiCE paper; the sparse CBM intervention is
the closest direct analogue of concept-space editing.

\section{Implementation in This Repository}

The original SpLiCE package is kept in \texttt{splice/}. The project-specific
integration is implemented mainly in \texttt{spur\_splice.py},
\texttt{splice/ssl\_regularization.py}, and the Waterbirds dataset utilities in
\texttt{experiments/spurious\_eval/}. The entry point \texttt{spur\_splice.py}
runs SimCLR training, checkpointing, optional periodic linear probing, and
logging. SpLiCE is enabled through \texttt{--splice\_mode}, with supported modes
\texttt{none}, \texttt{augment}, \texttt{corr\_reg}, and
\texttt{augment\_corr\_reg}.

\subsection{Workflow}

\begin{enumerate}

    \item If a SpLiCE mode is enabled, the training split is decomposed once by
    a frozen SpLiCE model. Concepts are ranked using target- and spurious-group
    metadata, and the shared dataset-level top-10 is cached. For the current
    Waterbirds runs this set is \texttt{bamboo, forests, whale, rainforest,
    raven, forest, seascape, lake, snorkeling, pond}.

    \item The Waterbirds SSL dataset wraps the usual WILDS-style subset with
    the cached selected-concept vectors. Each item returns two SimCLR views,
    the label, metadata, and a vector $c(x)\in\mathbb{R}^{10}_{\geq 0}$.

    \item In \texttt{augment} mode, $r(x)=\max_k c_k(x)$ controls routing. The
    threshold is a quantile computed on the training split, not a transferable
    absolute score. A high-score image produces the pair
    $[T_{\mathrm{std}}(x),T_{\mathrm{strong}}(x)]$; a low-score image produces
    two standard views. This gives SimCLR an explicit invariance pair without
    corrupting both views with the strong transform.

    \item In \texttt{corr\_reg} mode, the loss penalizes the complete matrix of
    correlations between SSL dimensions and the ten selected concepts. Both
    representations are centred within each target class before correlations
    are computed. This label-conditioned formulation avoids penalizing useful
    target information merely because target and background are correlated in
    the biased training split.

    \item Linear probing is decoupled from checkpoint retention and is run every
    25 SSL epochs during validation development. The target probe is trained on
    the balanced \texttt{ds\_train} split and evaluated on \texttt{val}. An
    optional second probe measures spurious-attribute leakage. A guarded
    \texttt{--final\_test} mode performs one final test probe only after model
    and hyperparameters have been fixed.
\end{enumerate}

\section{Supporting Tools}

Utility scripts and caches were added to make the SpLiCE configuration reproducible.
\texttt{tools/
discover\_splice\_spurious\_concepts.py} decomposes images by
metadata group and ranks concepts by how strongly they vary with the spurious
attribute while penalizing class-label effects. This helps choose candidate
background concepts. \texttt{tools/summarize\_splice\_scores.py} reports the
distribution of selected concept scores. Array jobs now use one file lock and
one discovery directory, so all tasks in a submission share exactly the same
concept set and cached vectors. A sparse SpLiCE-CBM entry point and optional
representation/probe interventions were also added. CelebA and SpurCIFAR10
dataset handling, metadata logging, evaluation guards, and tests were updated
as part of the same revision (commits \texttt{0198e3b}--\texttt{868092b}).

\section{Experiment Plan}

The development comparison uses seeds 0, 1, and 2. The augmentation sweep varies
the train-score quantile $q$; the regularization sweep varies $\lambda$. The
augmentation-component array keeps $q=0.75$ and changes only the strong
transform. Runs are compared using the mean of the final ten periodic validation
probes at SSL epoch 1000. Per-run peak epochs are not used for selection because
the WG curves are noisy. The final test will use one predeclared configuration
selected from aggregated validation results.

\begin{table}[h]
\centering
\caption{Fixed experiment settings}
\begin{tabularx}{\textwidth}{>{\bfseries}l X}
\toprule
Setting & Value \\
\midrule
Dataset & Waterbirds \\
SSL method & SimCLR \\
SSL encoder & \texttt{ResNet18-large} \\
SpLiCE model & \texttt{open\_clip:ViT-B-32}, \texttt{laion2b\_s34b\_b79k} \\
Vocabulary & \texttt{LAION}, size 10\,000 \\
Selected Top-10 SpLiCE concepts & bamboo,forests,whale,rainforest,raven,forest,seascape,
lake,snorkeling,pond \\
SpLiCE routing & maximum over selected vector; train quantile $q$ \\
Linear-probe train split & \texttt{ds\_train} (group-balanced) \\
Batch size & 256 \\
Learning rate & 0.01 \\
Epochs & 1000 \\
Weight decay & 1e-4 \\
Temperature & 0.05 \\
Linear probing & 100 epochs, every 25 SSL epochs \\
Development evaluation split & validation \\
Final evaluation split & test, only with \texttt{--final\_test} \\
\bottomrule
\end{tabularx}
\end{table}

\section{Current Validation Results (Automatic Pipeline)}

The W\&B project \texttt{gsgrechkin-rptu/Spur\_SpLiCE} was queried on
18 July 2026. Starting with the first relevant submission on 13 July at
19:11 UTC, 28 finished Waterbirds runs were matched to the two Slurm arrays.
The grouping was reconstructed from creation time, Git commit, seed, SpLiCE
mode, quantile or regularization weight, and strong-augmentation configuration.

\begin{table}[ht]
\centering
\small
\caption{Provenance groups for the current Waterbirds experiments.}
\label{tab:current_provenance}
\begin{tabularx}{\textwidth}{llllX}
\toprule
Array & Seed & Started (UTC) & Commit & W\&B run IDs \\
\midrule
Augmentation components & 2 & 13 Jul 19:11 & \texttt{272ee42} &
\texttt{dbn4o7gb, 0ptqafom, 6zpr2zbw, euvfljc2, mw3q4416} \\
Hyperparameter sweep & 2 & 14 Jul 09:08 & \texttt{868092b} &
\texttt{nc89rx14, uxenqlkt, fvtol29m, 5b4cvrig, q3o1ifpy, 2u5awr5z, apjwtvyn, 492n5krr, a9j4enb4} \\
Augmentation components & 0 & 15 Jul 19:41 & \texttt{868092b} &
\texttt{0t9p97ag, wyolxkon, 87ry202v, ogz7t6r4, g77u3g1z} \\
Hyperparameter sweep & 0 & 17 Jul 12:12 & \texttt{868092b} &
\texttt{8cfxm2q2, si31b8ng, tpvleejg, 33ko4h4h, 7nz2e6o2, g4jel2bp, ftgmmifc, m6bgwc23, naet2tg6} \\
\bottomrule
\end{tabularx}
\end{table}

\subsection{Hyperparameter Sweep}

Table~\ref{tab:current_sweep} reports the W\&B summary fields
\texttt{Average over last 10 linear val acc} and
\texttt{Average over last 10 linear val worst-group acc}. Seed 1 has not yet
been run under the revised protocol and is deliberately left as a TODO.

\begin{table}[ht]
\centering
\small
\caption{Current validation sweep. Values are average / worst-group accuracy (\%).}
\label{tab:current_sweep}
\begin{tabular}{llccc}
\toprule
Method & Hyperparameter & Seed 0 & Seed 1 & Seed 2 \\
\midrule
SimCLR & -- & 49.27 / 41.93 & \texttt{TODO} & 50.60 / 37.69 \\
SpLiCE-AUG & $q=0.50$ & 49.83 / 42.96 & \texttt{TODO} & 53.64 / 46.96 \\
SpLiCE-AUG & $q=0.75$ & 52.45 / 43.11 & \texttt{TODO} & 56.80 / 49.62 \\
SpLiCE-AUG & $q=0.90$ & 53.52 / 47.25 & \texttt{TODO} & 53.02 / 41.99 \\
SpLiCE-AUG & $q=0.95$ & 50.59 / 44.64 & \texttt{TODO} & 53.08 / 43.83 \\
SpLiCE-REG & $\lambda=0.001$ & 55.50 / 53.13 & \texttt{TODO} & 52.31 / 41.07 \\
SpLiCE-REG & $\lambda=0.01$ & 49.22 / 39.61 & \texttt{TODO} & 50.41 / 40.86 \\
SpLiCE-REG & $\lambda=0.1$ & 47.52 / 39.70 & \texttt{TODO} & 53.71 / 43.64 \\
SpLiCE-REG & $\lambda=1.0$ & 51.88 / 44.64 & \texttt{TODO} & 50.97 / 36.96 \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\small
\caption{Aggregate current sweep over the two available seeds. The standard
deviation is the sample standard deviation; $\Delta$WG is relative to the
matched two-seed baseline.}
\label{tab:current_sweep_aggregate}
\begin{tabular}{lcccc}
\toprule
Configuration & $n$ & Avg. Acc. & WG Acc. & $\Delta$WG \\
\midrule
SimCLR & 2 & $49.94\pm0.94$ & $39.81\pm3.00$ & -- \\
SpLiCE-AUG $q=0.50$ & 2 & $51.73\pm2.69$ & $44.96\pm2.83$ & $+5.15$ \\
SpLiCE-AUG $q=0.75$ & 2 & $54.62\pm3.07$ & $46.37\pm4.61$ & $+6.56$ \\
SpLiCE-AUG $q=0.90$ & 2 & $53.27\pm0.35$ & $44.62\pm3.72$ & $+4.81$ \\
SpLiCE-AUG $q=0.95$ & 2 & $51.83\pm1.76$ & $44.23\pm0.57$ & $+4.43$ \\
SpLiCE-REG $\lambda=0.001$ & 2 & $53.91\pm2.26$ & $47.10\pm8.52$ & $+7.29$ \\
SpLiCE-REG $\lambda=0.01$ & 2 & $49.81\pm0.84$ & $40.24\pm0.88$ & $+0.43$ \\
SpLiCE-REG $\lambda=0.1$ & 2 & $50.62\pm4.38$ & $41.67\pm2.79$ & $+1.86$ \\
SpLiCE-REG $\lambda=1.0$ & 2 & $51.43\pm0.65$ & $40.80\pm5.43$ & $+0.99$ \\
\bottomrule
\end{tabular}
\end{table}

All four augmentation quantiles improve WG accuracy in the two-seed mean, which
is encouraging for the main hypothesis. The paired improvements are especially
large for seed 2, however: for $q=0.75$ they are $+1.18$ and $+11.94$ points for
seeds 0 and 2 respectively. Thus two seeds are not sufficient evidence of a
stable effect. The largest mean WG gain is obtained by
$\lambda=0.001$, but its WG standard deviation is also the largest. Among the
augmentation settings, $q=0.90$ has the most consistent paired WG improvement
($+5.33$ and $+4.30$), whereas $q=0.75$ has the best mean average and WG
accuracy. Seed 1 and additional confirmation seeds must be completed before
choosing between these candidates.

\subsection{Strong-Augmentation Components}

The component array is a separate ablation family: it enables periodic rank
evaluation and the spurious leakage probe, while the hyperparameter sweep does
not. It should be used to compare augmentation components within the family,
not pooled with the sweep as exact replicates.

\begin{table}[ht]
\centering
\small
\caption{Current $q=0.75$ augmentation ablation. Values are average / WG
validation accuracy (\%).}
\label{tab:current_augmentations}
\begin{tabular}{lccc}
\toprule
Strong transform & Seed 0 & Seed 1 & Seed 2 \\
\midrule
All components & 50.33 / 39.29 & \texttt{TODO} & 56.16 / 49.36 \\
Crop only & 52.44 / 44.80 & \texttt{TODO} & 55.40 / 46.10 \\
Color jitter only & 48.23 / 39.08 & \texttt{TODO} & 54.96 / 44.84 \\
Grayscale only & 53.53 / 48.52 & \texttt{TODO} & 52.70 / 38.80 \\
Gaussian blur only & 44.79 / 34.12 & \texttt{TODO} & 52.49 / 44.35 \\
\bottomrule
\end{tabular}
\end{table}

Across the two available seeds, crop is currently the strongest individual
component ($45.45\pm0.92$ WG), while using all components gives
$44.32\pm7.12$ WG. Grayscale reverses strongly between seeds and blur is the
weakest mean setting. Consequently, the data do not support the earlier claim
that all four augmentations are uniformly best; the interaction with seed is
substantial and seed 1 is essential.

The leakage probe remains above 91\% average validation accuracy for every
component run (approximately 91.2--92.4\%). This does not yet support the
stronger claim that the learned representation has stopped encoding the
background. It is possible for target WG accuracy to improve while background
information remains linearly accessible, so both metrics should continue to be
reported.

\subsection{Protocol Caveat and Reproducibility Repair}

Rank evaluation was intended to be observational, but the current
\texttt{extract\_normalized\_train\_features} function iterates the same shuffled,
augmented training loader used by SSL. This can advance loader and worker random
state and thereby change later training batches. W\&B provides a useful matched
check: runs \texttt{tpvleejg} and \texttt{0t9p97ag} have the same seed, commit,
model name, concepts, $q=0.75$, and augmentation policy. Their meaningful config
differences are \texttt{rank\_eval\_freq=0} versus 100 and the additional leakage
probe. Nevertheless their final-ten WG values are 43.11 and 39.29. This does not
prove that rank evaluation caused the entire difference, but it demonstrates
that the historical two families are not exchangeable replicates.

The implementation has now been repaired for future runs. Rank diagnostics use
a dedicated non-augmented, non-shuffled train loader with the deterministic
evaluation transform and an independent generator. Periodic probes preserve
Python, NumPy, CPU Torch, and CUDA RNG states. Persistent checkpoints save and
restore those states, the SSL DataLoader generator, optimizer, and AMP scaler,
so resumed jobs continue the same random sequence. Frozen SpLiCE/CLIP loading
and W\&B initialization are also RNG-isolated so paired methods start from the
same SSL initialization. The resolved numeric routing
threshold is now written to \texttt{args.json} and W\&B as
\texttt{splice\_score\_threshold\_resolved}. These changes make diagnostics
observational, but they cannot retroactively repair the July 13--17 runs; the
selected settings must be rerun with the repaired protocol.

\subsection{Predeclared Final-Test Table}

\begin{table}[ht]
\centering
\caption{Final Waterbirds test results. Do not fill until validation selection
and protocol repair are complete.}
\label{tab:final_test_todo}
\begin{tabular}{lccc}
\toprule
Method & Selected hyperparameter & Avg. test acc. & WG test acc. \\
\midrule
SimCLR baseline & -- & \texttt{TODO} & \texttt{TODO} \\
Automatic SpLiCE-AUG & $q=\texttt{TODO}$ & \texttt{TODO} & \texttt{TODO} \\
Conditional SpLiCE-REG & $\lambda=\texttt{TODO}$ & \texttt{TODO} & \texttt{TODO} \\
Combined AUG+REG & $q,\lambda=\texttt{TODO}$ & \texttt{TODO} & \texttt{TODO} \\
Sparse SpLiCE-CBM intervention & intervention=\texttt{TODO} & \texttt{TODO} & \texttt{TODO} \\
\bottomrule
\end{tabular}
\end{table}

\FloatBarrier

\section{Preliminary Legacy Results (Scalar Control)}

The following tables were produced with the implementation preceding commit
\texttt{0198e3b}. They used manual absolute thresholds, a scalar correlation
control, and a different evaluation protocol. They are retained for historical
context and must not be pooled statistically with Tables~\ref{tab:current_sweep}
and~\ref{tab:current_augmentations}.

\begin{longtable}{>{\raggedright\arraybackslash}p{3.0cm}
                  >{\centering\arraybackslash}p{1.5cm}
                  >{\centering\arraybackslash}p{1.5cm}
                  >{\centering\arraybackslash}p{2cm}
                  >{\centering\arraybackslash}p{2cm}}
\label{tab:results}\\
\toprule
Method & Threshold & Weight & Avg. Acc. & WG Acc. \\
\toprule
\endhead
\multicolumn{5}{l}{\textbf{Seed 0}} \\
\toprule
Baseline SSL & -- & -- & 47.13 & 38.86 \\
SpLiCE-AUG & \textbf{0.03} & -- & \textbf{56.32} & \textbf{52.66} \\
SpLiCE-AUG & 0.05 & -- & 53.67 & 51.31 \\
SpLiCE-AUG & 0.085 & -- & 51.27 & 46.75 \\
SpLiCE-AUG & 0.1 & -- & 49.97 & 43.89 \\
\midrule
SpLiCE-REG & -- & 0.0001 & 49.77 & 42.99 \\
SpLiCE-REG & -- & 0.0005 & 51.69 & 49.70 \\
SpLiCE-REG & -- & 0.001 & 48.96 & 41.17 \\
SpLiCE-REG & -- & 0.005 & 52.30 & 47.95 \\
\toprule
\multicolumn{5}{l}{\textbf{Seed 1}} \\
\toprule
Baseline SSL & -- & -- & 57.75 & 49.51 \\
SpLiCE-AUG & \textbf{0.03} & -- & \textbf{58.12} & \textbf{52.46} \\
SpLiCE-AUG & 0.05 & -- & 54.54 & 51.52 \\
SpLiCE-AUG & 0.085 & -- & 54.75 & 49.02 \\
SpLiCE-AUG & 0.1 & -- & 56.30 & 53.57 \\
\midrule
SpLiCE-REG & -- & 0.0001 & 48.22 & 41.07 \\
SpLiCE-REG & -- & 0.0005 & 56.36 & 51.78 \\
SpLiCE-REG & -- & 0.001 & 57.93 & 51.76 \\
SpLiCE-REG & -- & 0.005 & 52.36 & 45.96 \\
\toprule
\multicolumn{5}{l}{\textbf{Seed 2}} \\
\toprule
Baseline SSL & -- & -- & 52.02 & 44.98 \\
SpLiCE-AUG & 0.03 & -- & 55.47 & 49.12 \\
SpLiCE-AUG & 0.05 & -- & 52.78 & 40.21 \\
SpLiCE-AUG & \textbf{0.085} & -- & \textbf{55.40} & \textbf{50.46} \\
SpLiCE-AUG & 0.1 & -- & 53.94 & 43.52 \\
\midrule
SpLiCE-REG & -- & 0.0001 & 50.90 & 40.53 \\
SpLiCE-REG & -- & 0.0005 & 51.32 & 37.32 \\
SpLiCE-REG & -- & 0.001 & 55.56 & 51.15 \\
SpLiCE-REG & -- & 0.005 & 51.93 & 41.71 \\
\bottomrule
\end{longtable}

Additional legacy augmentation-specific tests were used to identify which
transformations might contribute to SpLiCE-guided SSL. Their ranking is
seed-dependent: all components maximize average accuracy for seed 1, while
grayscale gives the highest seed-1 WG accuracy and crop is strongest for seed 0.
These exploratory results motivated the revised component array but do not
establish a universally best augmentation policy.

\begin{table}[ht]
\centering
\caption{Effect of individual augmentations on linear evaluation accuracy for \textit{waterbirds} dataset}
\label{tab:individual_augmentations_waterbirds}
\begin{tabular*}{0.75\textwidth}{@{\extracolsep{\fill}}lcc}
\toprule
Augmentation & Avg. Acc. & WG Acc. \\
\midrule
\multicolumn{3}{l}{\textbf{Seed 0}} \\
Crop & \textbf{56.08} & 49.01 \\
Color jitter & 54.73 & 48.29 \\
Grayscale & 51.56 & 47.20 \\
Gaussian blur & 51.30 & 44.37 \\
All aug. & 55.28 & \textbf{49.83} \\
\midrule
\multicolumn{3}{l}{\textbf{Seed 1}} \\
Crop & 53.86 & 49.97 \\
Color jitter & 55.47 & 49.12 \\
Grayscale & 57.27 & 55.98 \\
Gaussian blur & 50.47 & 43.56 \\
All aug. & 58.12 & 52.46 \\
\midrule
\multicolumn{3}{l}{\textbf{Seed 2}} \\
Crop & 54.12 & 39.54 \\
Color jitter & 52.38 & 45.78 \\
Gaussian blur & 53.85 & 45.91 \\
Grayscale & 49.53 & 40.11 \\
All aug. & 55.47 & 49.12 \\
\bottomrule
\end{tabular*}
\end{table}

\pagebreak
A short overview of used augmenation parameters. For each experiment, only one augmentation was  replaced by its strong variant, while all remaining augmentations were kept identical to the standard SimCLR pipeline. The GaussianBlur experiment evaluates the effect of enabling the Gaussian blur transform rather than changing its internal parameters.

\begin{table}[ht]
\centering
\caption{Strong augmentation variants evaluated in the SpLiCE experiments.}
\label{tab:aug_settings}
\begin{tabular}{lll}
\toprule
Augmentation & Standard SimCLR & Strong variant \\
\midrule
Crop & scale $(0.2,1.0)$ & scale $(0.08,1.0)$ \\
ColorJitter & $(0.4,0.4,0.4,0.1),\;p=0.8$ &
$(0.8,0.8,0.8,0.2),\;p=0.9$ \\
RandomGrayscale & $p=0.2$ & $p=0.3$ \\
GaussianBlur & disabled & enabled ($p=0.5,\;k=23,\;\sigma=(0.1,2.0)$) \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Effect of individual augmentations on linear evaluation accuracy for \textit{SpurCIFAR10} dataset}
\label{tab:individual_augmentations_spurcifar10}
\begin{tabular*}{0.75\textwidth}{@{\extracolsep{\fill}}lcc}
\toprule
Augmentation & Avg. Acc. & WG Acc. \\
\midrule
\multicolumn{3}{l}{\textbf{Seed 0}} \\
Crop & 62.06 & 30.35 \\
Color jitter & 61.39 & 22.77 \\
Grayscale & 62.06 & 26.58 \\
Gaussian blur & 63.42 & 25.69 \\
All aug. & 63.44 & 28.64 \\
\midrule
\multicolumn{3}{l}{\textbf{Seed 1}} \\
Crop & 63.09 & 32.55 \\
Color jitter & 61.50 & 29.06 \\
Grayscale & 62.85 & 31.54 \\
Gaussian blur & 62.36 & 32.58 \\
All aug. & 63.67 & 34.83 \\
\midrule
\multicolumn{3}{l}{\textbf{Seed 2}} \\
Crop & 62.70 & 27.82 \\
Color jitter & 62.69 & 37.80 \\
Gaussian blur & 62.09 & 21.30 \\
Grayscale & 62.62 & 20.86 \\
All aug. & 63.23 & 27.79 \\
\bottomrule
\end{tabular*}
\end{table}

\FloatBarrier

\section{SpLiCE Concept Scoring and Regularization Formulas}

Let \(c_k(x)\) denote the SpLiCE weight of concept \(k\) for image \(x\).
For Waterbirds, each image has a target label $ y \in \{0,1\}$
and a spurious/background attribute $ s \in \{0,1\}. $
For each concept \(k\), the discovery script first computes the mean SpLiCE concept weight inside each metadata group:
\[
\mu_k(s,y)
=
\mathbb{E}\left[c_k(x)\mid s(x)=s,\ y(x)=y\right].
\]
Thus, for the binary Waterbirds setting, each concept has four group means:
\[
\mu_k(0,0),\quad
\mu_k(1,0),\quad
\mu_k(0,1),\quad
\mu_k(1,1).
\]

\subsection{Spurious Effect}

The spurious effect measures how much concept \texttt{k} changes when the background \texttt{s} changes, while keeping the bird label \texttt{y} fixed:
\[
\Delta^{\mathrm{spur}}_k(y)
=
\mu_k(1,y)-\mu_k(0,y).
\]
The reported spurious value is the average absolute effect across target classes:
\[
\mathrm{Spurious}_k
=
\frac{1}{2}
\left(
\left|\mu_k(1,0)-\mu_k(0,0)\right|
+
\left|\mu_k(1,1)-\mu_k(0,1)\right|
\right).
\]
A high spurious value means that the concept is strongly related to spurious feature, which is a desirable behavior in our case.
\subsection{Target Effect}

The target effect measures how much concept \texttt{k} changes when the bird label \texttt{y} changes, while keeping the background \texttt{s} fixed:
\[
\Delta^{\mathrm{target}}_k(s)
=
\mu_k(s,1)-\mu_k(s,0).
\]
The reported target value is:
\[
\mathrm{Target}_k
=
\frac{1}{2}
\left(
\left|\mu_k(0,1)-\mu_k(0,0)\right|
+
\left|\mu_k(1,1)-\mu_k(1,0)\right|
\right).
\]
A high target value means that the concept is strongly related to the actual bird class, which is not desirable when searching for background-related concepts.

\subsection{Instability}

The instability term measures whether the \emph{magnitude} of the spurious
effect is consistent across target classes. Define
\[
 a_k = \left|\mu_k(1,0)-\mu_k(0,0)\right|,
\]
\[
 b_k = \left|\mu_k(1,1)-\mu_k(0,1)\right|.
\]
The script computes the standard deviation of these two values:
\[
\mathrm{Instability}_k
=
\sqrt{
\frac{1}{2}
\left[
(a_k-\bar{d}_k)^2
+
(b_k-\bar{d}_k)^2
\right]
},
\]
where
\[
\bar{d}_k
=
\frac{a_k+b_k}{2}.
\]
Low instability means that the concept behaves similarly for both bird classes. High instability means that the concept only behaves like a spurious concept for one target class.

\subsection{Final Discovery Score}

The final concept discovery score is computed as
\[
\mathrm{Score}_k
=
\mathrm{Spurious}_k
-
\lambda_y \mathrm{Target}_k
-
\lambda_{\mathrm{inst}} \mathrm{Instability}_k.
\]
Here, \(\lambda_y\) is controlled by \texttt{--label\_penalty}, and
\(\lambda_{\mathrm{inst}}\) is controlled by \texttt{--instability\_penalty}.
By default, both values are equal to \(1.0\), so the default formula is
\[
\mathrm{Score}_k
=
\mathrm{Spurious}_k
-
\mathrm{Target}_k
-
\mathrm{Instability}_k.
\]
Therefore, the script ranks concepts highly when they have a strong spurious effect, a weak target effect, and low instability.

\subsection{Correlation Regularization Penalty}

In the revised \texttt{corr\_reg} mode, SpLiCE retains the selected concept
vector $c_i\in\mathbb{R}^{K}_{\geq0}$ for each training image, with $K=10$ in
the current experiments. The SSL encoder produces
\[
h_i = f_{\theta}(x_i) \in \mathbb{R}^d.
\]
Because background and target are correlated in the biased training set, the
default loss first centres embeddings and concept vectors separately within
each target class $B_y=\{i:y_i=y\}$:
\[
\widetilde h_i=h_i-\frac{1}{|B_{y_i}|}\sum_{j\in B_{y_i}}h_j,
\qquad
\widetilde c_i=c_i-\frac{1}{|B_{y_i}|}\sum_{j\in B_{y_i}}c_j.
\]
It then computes every embedding--concept correlation
\[
R_{jk}=
\frac{\sum_i \widetilde h_{ij}\widetilde c_{ik}}
{\sqrt{\sum_i\widetilde h_{ij}^{2}}
 \sqrt{\sum_i\widetilde c_{ik}^{2}}+\epsilon}.
\]
The revised vector-valued penalty is
\[
\mathcal{L}_{\mathrm{SpLiCE}}
=
\lambda_{\mathrm{SpLiCE}}
\frac{1}{dK}
\sum_{j=1}^{d}\sum_{k=1}^{K}R_{jk}^2.
\]
Here, \(\lambda_{\mathrm{SpLiCE}}\) is the regularization weight controlled by
\texttt{--splice\_weight}. The full SSL training objective becomes
\[
\mathcal{L}
=
\mathcal{L}_{\mathrm{SimCLR}}
+
\mathcal{L}_{\mathrm{SpLiCE}}.
\]
Thus the regularizer discourages any linear SSL feature from encoding any of
the selected SpLiCE coordinates \emph{within a fixed target class}. It does not
identify a one-to-one ``background neuron'', and because it uses target labels
inside the SSL loss it must be described as label-conditioned rather than fully
label-free.

\subsection{Score Distribution in Concept Summarization}

After concept discovery, we select a set of concepts \(C\), for example
\[
C =
\{
\texttt{bamboo},
\texttt{forests},
\texttt{whale},
\texttt{rainforest},
\texttt{raven},
\texttt{forest},
\texttt{seascape},
\texttt{lake},
\texttt{snorkeling},
\texttt{pond}
\}.
\]
For every image \(x_i\), the summarization script computes one scalar score
\(r_i\) from the selected concept weights.

If \texttt{--splice\_score\_reduction max} is used, then
\[
r_i
=
\max_{k \in C} c_k(x_i).
\]
The score distribution is the empirical distribution of these image-level
scores:
\[
\{r_1,r_2,\ldots,r_n\}.
\]
The reported values such as the median, \(p75\), \(p90\), and \(p95\) are
percentiles of this distribution.

For a threshold \(\tau\), the script reports the number of images whose score is
above the threshold:
\[
N_{\tau}
=
\sum_{i=1}^{n}
\mathbf{1}[r_i \geq \tau],
\]
and the corresponding fraction:
\[
F_{\tau}
=
\frac{1}{n}
\sum_{i=1}^{n}
\mathbf{1}[r_i \geq \tau].
\]
This tells us how many training images will be treated as high SpLiCE-score
examples. In the augmentation method, images with
\[
r_i \geq \tau
\]
receive the stronger augmentation transform.


\section{Reproduction and Next Experiment Commands}

\paragraph{1. Standalone concept discovery (optional audit).}
The Slurm arrays perform this step automatically and share its output across
tasks. The standalone command is useful for auditing the selected concepts:

\begin{verbatim}
python tools/discover_splice_spurious_concepts.py \
  --dataset waterbirds \
  --data_folder ./datasets \
  --split train \
  --out_path outputs/waterbirds_splice_concepts.json \
  --top_k 10
\end{verbatim}

In one run, the top-10 concepts were:
\begin{verbatim}
8305 bamboo      score=0.010589 spurious=0.016997 target=0.003204 instability=0.003204
2908 forests     score=0.004767 spurious=0.006347 target=0.000822 instability=0.000758
6265 whale       score=0.001928 spurious=0.002307 target=0.000189 instability=0.000189
4730 rainforest  score=0.001892 spurious=0.003017 target=0.000565 instability=0.000560
591  raven       score=0.001616 spurious=0.002540 target=0.000715 instability=0.000209
9479 forest      score=0.001492 spurious=0.002912 target=0.000710 instability=0.000710
4623 seascape    score=0.001484 spurious=0.002110 target=0.000313 instability=0.000313
9411 lake        score=0.001310 spurious=0.001343 target=0.000017 instability=0.000017
1158 snorkeling  score=0.001178 spurious=0.001433 target=0.000143 instability=0.000112
8045 pond        score=0.001129 spurious=0.001685 target=0.000278 instability=0.000278
\end{verbatim}

Here, \texttt{spurious} measures how much the concept changes with the
background attribute after conditioning on the bird label. \texttt{target}
measures how much the same concept changes with the bird label after
conditioning on the background. \texttt{instability} measures whether the
spurious effect is inconsistent across target classes. The final
\texttt{score} is the ranking value used by the script: high spurious effect is
rewarded, while target-label effect and instability are penalized. Therefore,
higher-score concepts are better candidates for representing the spurious
background signal.

\paragraph{2. Summarize selected concept scores.}
After selecting candidate concepts, the next step is to check the distribution
of their SpLiCE scores across the training split. This helps choose a threshold
for the augmentation method.

\begin{verbatim}
python tools/summarize_splice_scores.py \
  --data_folder ./datasets \
  --split train \
  --splice_concepts "bamboo,forests,whale,rainforest,raven,forest,seascape,lake,
                     snorkeling,pond" \
  --splice_score_reduction max \
  --out_path outputs/waterbirds_splice_score_summary.json
\end{verbatim}

The output first resolves each concept to its vocabulary index, then reports the
score distribution over all images in the split. The percentiles summarize how
large the selected-concept score usually is, while the threshold counts show how
many images would be treated as high-score examples for each candidate
threshold. For example, in the shown run, threshold 0.085 selects 1007 training
images, or about 21.0\% of the split.

\paragraph{3. Complete the missing seed.}
With the repaired rank-evaluation loader, complete seed 1 for both current
arrays. Environment-variable names are case-sensitive: the scripts read
\texttt{SEED}, not \texttt{seed}.

\begin{verbatim}
sbatch --export=ALL,SEED=1 --array=0-8 \
  waterbirds_SpLiCE_hyperparameter_array.sbatch

sbatch --export=ALL,SEED=1 --array=0-4 \
  waterbirds_Augmentation_array.sbatch
\end{verbatim}

\paragraph{4. Confirm only the validation shortlist on additional seeds.}
Array indices 0, 2, 3, and 5 correspond to baseline, AUG $q=0.75$, AUG
$q=0.90$, and REG $\lambda=0.001$. After seed 1, these are the current
shortlist for seeds 3 and 4:

\begin{verbatim}
sbatch --export=ALL,SEED=3 --array=0,2,3,5 \
  waterbirds_SpLiCE_hyperparameter_array.sbatch
sbatch --export=ALL,SEED=4 --array=0,2,3,5 \
  waterbirds_SpLiCE_hyperparameter_array.sbatch
\end{verbatim}

\paragraph{5. Transfer the protocol to SpurCIFAR10.}
The same arrays now accept \texttt{DATASET} as an exported variable and choose
\texttt{resnet18} automatically for 32$\times$32 inputs. Dataset-level concept
discovery and caches remain separate because the dataset name is included in
their paths and fingerprints.

\begin{verbatim}
sbatch --job-name=SpurCIFAR10_SpLiCE_sweep \
  --export=ALL,DATASET=spur_cifar10,SEED=0 \
  waterbirds_SpLiCE_hyperparameter_array.sbatch

sbatch --job-name=SpurCIFAR10_SpLiCE_aug \
  --export=ALL,DATASET=spur_cifar10,SEED=0 \
  waterbirds_Augmentation_array.sbatch
\end{verbatim}

The existing case statement fixes $q$ and $\lambda$ for each array index, so
exporting \texttt{SPLICE\_SCORE\_QUANTILE} or \texttt{SPLICE\_WEIGHT} currently
does not override them. Finer sweeps require new case entries or changing those
assignments to environment-default form. The most useful controls after the
replication seeds are: (i) global strong augmentation applied to a random or all
images, to distinguish SpLiCE routing from augmentation strength; (ii) shuffled
SpLiCE scores, to test whether semantic targeting matters; (iii) conditional
versus unconditional REG; (iv) combined AUG+REG; and (v) sparse SpLiCE-CBM
representation/probe interventions. Final test jobs should be launched only
after this validation stage and must use the explicit \texttt{--final\_test}
guard.

\end{document}
